This CapstoneMilestoneReport repository is based on Lalit A Patel's Milestone Report for Coursera DSS Capstone Project. Html version of this Milestone Report has been posted at http://rpubs.com/Lapyl/70623 .

This Milestone Report is not outstanding or exemplary. Its R script is not satisfactory. However, its html formattintg features may be of some use elsewhere.

## Features

This Milestone Report's html file was prepared by applying knit html on Rmd file in RStudio. The html file has the following 2 formattintg features.

##### Undecorated hyperlinks

* Copy "RStudio/resources/markdown.css" as "MilesLap.css" in the direcory of "MilesLap.Rmd".
* Modify "MileLap.css": Modify style for "a" to remove decoration on hyperlinks. Modify style for "code" to show R code chunks in a different color.
* Add the following after the top section in the Rmd file in RStudio.

```{r}
    <link rel=StyleSheet type="text/css" href="MilesLap.css" media=screen></link>
```

##### Controlled display for R code chunks

* Add the following after the above link tag in the Rmd file in RStudio.

```{r}
    <script language="JavaScript">
	function ShowHide(divId) {
	if(document.getElementById(divId).style.display == 'none')
	{document.getElementById(divId).style.display='block';}
	else {document.getElementById(divId).style.display = 'none';}
	}
	</script>
```

* Add the following before each R code chunk in the Rmd file in RStudio.

```{r}
	<a onclick ="javascript:ShowHide('SomeDivName')" href="javascript:;">
    	<font color="green">SomeCodeChunkName * Show/Hide</font></a>
	<div class="mid" id="SomeDivName" style="DISPLAY: none">
```

* Add the following after each R code chunk in the Rmd file in RStudio.
```{r}
	</div>
```

## MilesLap.Rmd

MilesLap.Rmd is R markdown file written in RStudio. Clicking on MilesLap.Rmd in this repository displays its formatted content. Its raw code can be viewed as follows.

##### Raw Rmd by URL

* Write "https://raw.githubusercontent.com/Lapyl/CapstoneMilestoneReport/master/MilesLap.Rmd" in the URL box of your Internet browser.

##### Raw Rmd by Button

* Click on MilesLap.Rmd in this repository. This will display its formatted content.
* Click on Raw button from the top bar. This will display its raw code.

## MilesLap.html

MilesLap.html is html file knited from RStudio. Clicking on MilesLap.html in this repository displays its source code. Its formatted content can be viewed as follows.

##### Formatted Html on Rpubs

* Write "http://rpubs.com/Lapyl/70623" in the URL box of your Internet browser.

##### Formatted Html by URL

* Write "http://htmlpreview.github.io/?https://github.com/Lapyl/CapstoneMilestoneReport/blob/master/MilesLap.html" in the URL box of your Internet browser.

##### Formatted Html by Dialog

* Write "http://htmlpreview.github.io" in the URL box of your Internet browser.
* Write "https://github.com/Lapyl/CapstoneMilestoneReport/blob/master/MilesLap.html" in the dialog box of the resulting web page.

##### Formatted Html by Button

* Click on MilesLap.html in this repository. This will display its source code.
* Click on the "Open this file in GitHub for Windows" icon next to the History button in the box above its source code. This will display its formatted content.